---
title: "GatherAndCleaningData Project"
author: "Sujata"
date: "January 8, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(data.table)
library(tidyr)
```

## R Markdown
###Get the data

Dowload the file and put in the folder and unzip

```{r download}
curPath <- file.path(getwd(), "./UCI HAR Dataset")

fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2fprojectfiles%2fUCI%20HAR%20Dataset.zip"
f <- "Dataset.zip"
if (!file.exists(curPath)) {
  dir.create(curPath)
}
if (!file.exists(file.path(curPath,f))) {
  download.file(fileUrl,file.path(curPath,f))
  
  unzip(file.path(curPath,f))
}
```

## Read the required dataset files except for the files in the "Intertia Signals"" folder.
Using "read.table" since it is easier to remove duplicated columns in data.frame.


```{r read_data, echo=FALSE}
uci_X_train <- read.table(file.path(curPath, "train", "X_train.txt"), header=FALSE)
#head(uci_X_train)

uci_X_test <- read.table(file.path(curPath, "test", "X_test.txt"), header=FALSE)
#head(uci_X_test)

activity_train <- fread(file.path(curPath, "train" ,"y_train.txt"))
#head(activity_train)

activity_test <- fread(file.path(curPath, "test", "y_test.txt"))
#head(activity_test)

subject_train <- fread(file.path(curPath, "train", "subject_train.txt"))
#head(subject_train)

subject_test <- fread(file.path(curPath, "test", "subject_test.txt"))
#head(subject_test)

```

#1. Merge training and the test datasets to create one dataset

```{r merge}

uci <- rbind(uci_X_train, uci_X_test)

activity <- rbind(activity_train, activity_test)

dtSubject <- rbind(subject_train, subject_test)

dtSubject <- setNames(dtSubject, "subject")

```

#2. Extracts only the measure ments on the mean and standard deviation for each measurement
a. Read feature labels
b. Set the feature labels on the UCI daataset
c. Select only mean() and std() measurements
```{r extract_fields}
features <- fread(file.path(curPath, "features.txt"))
#head(features)

pattern <- c("mean\\(\\)", "std\\(\\)")
matchExpression <- paste(pattern, collapse = "|")

uci <- setNames(uci, as.character(features$V2))

#removing duplicated names assuming this is one of cleanup activty bo stated but required to extract only #mean and std
uci <- uci[,!duplicated(names(uci))]
uci_subset <- select(uci, matches(matchExpression))
```
#3. Uses descriptive activity names to name of the activities in the dataset
a. Load Activity labels
b. join with activity data.table 
This will be used to add descriptive name to the activities
```{r activity, echo=FALSE}

activity_labels <- fread(file.path(curPath, ".","activity_labels.txt"))
#head(activity_labels)

dtActivity <- activity %>%
 left_join(activity_labels) %>%
 setNames(c("activityNum", "activityName"))

```
#4. Appropriately labels the data set with descriptive variable names
a. Merge Subject and Activity tables
b. Merge the above with UCI dataset
c. Melt the data table to reshape it from a short and wide format to a tall and narrow format.
d. Seperate featuers from feature Name using helper function grepVar

```{r labeling}

dtSubject <- cbind(dtSubject, dtActivity)
dt <- cbind(dtSubject, data.table(uci_subset))
setkey(dt, subject, activityNum, activityName)
dt <- melt(dt, id=c("subject","activityNum","activityName"), measure.vars=names(dt)[4:length(names(dt))])

grepVar <- function(regex1, regex2, name1, name2) {
  ifelse(grepl(regex1, dt$variable, ignore.case = T), name1, 
      ifelse(grepl(regex2, dt$variable, ignore.case = T), name2, "NA"))
}

dt$featureDomain <- grepVar("^t", "^f", "Time", "Frequency")
dt$featureSensor <- grepVar("Acc","Gyro","Accelerometer","Gyroscope")
dt$featureMotion <- grepVar("BodyAcc","GravityAcc","Body","Gravity")
dt$featureMeasure <- grepVar("mean()","std()","Mean","SD")
dt$featureJerk <- ifelse(grepl("Jerk", dt$variable, ignore.case = T), "Jerk", "NA")
dt$featureMagnitude <- ifelse(grepl("Mag", dt$variable, ignore.case = T), "Magnitude", "NA")
dt$featureAxis <- ifelse(grepl("-X", dt$variable, ignore.case = T), "X", 
         ifelse(grepl("-Y", dt$variable, ignore.case = T), "Y", 
         ifelse(grepl("-Z", dt$variable, ignore.case = T), "Z",      
                "Other")))
```
#5. Create a second, independent tidy data set with average of each variable for each activity and each subject
a. Create Tidydataset
b. save Tidydataset

```{r average_activity_subject}

#Ideally it may makes sense to convert activityName, featureDomain, featureSensor, featureMotion, 
#       featureJerk, featureMagnitude, featureMeasure, featureAxis as factors

setkey(dt, subject, activityNum, activityName, featureDomain, featureSensor, featureMotion, 
       featureJerk, featureMagnitude, featureMeasure, featureAxis)

dtTidySet <- dt[, list(count=.N, average = mean(value)), by=key(dt)]

write.table(dtTidySet, "TidyDataset.txt", sep="\t")

```

